---
title: "Business Intelligence Sample model Markdown"
author: "Nicholas Munene"
date: "21/6/2024"
output:
  github_document: 
    toc: yes
    toc_depth: 4
    fig_width: 6
    fig_height: 4
    df_print: default
editor_options:
  chunk_output_type: console
---

# MY Details

+---------------------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **Name**                                         |
|                                                   |                                                                                                                                                                          |
|                                                   |                                                                                                                                                                        |
|                                                   |                                                                                                                                                                          |
|                                                   |      Nicholas Munene                                                                                                                                       |
+---------------------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **GitHub repo**                                   | Nick's portfolio                                                                                                      |
+---------------------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **Course Done**                                   | Bachelors in Business  Information Technology                                                                                                                                                                 |
+---------------------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **Specialization**                                | Business Intelligence II                                                                                                                                                 |
+---------------------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **Program**                                       | Bachelor of Business Information Technology                                                                                                                              |
+---------------------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **Course   Duration**                             | April 2020 - June 2024                                                                                                                             |
+---------------------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------+

# Setup Chunk
In this markdown I will be creating a model that analyses data fro the breast cancer dataset by following the stages of data analysis.
We start by installing all the required packages We start by installing all the required packages
```{r}


if (require("languageserver")) {
  require("languageserver")
} else {
  install.packages("languageserver", dependencies = TRUE,
                   repos = "https://cloud.r-project.org")
}



# STEP 1. Install and Load the Required Packages ----
## caret ----
if (require("caret")) {
  require("caret")
} else {
  install.packages("caret", dependencies = TRUE,
                   repos = "https://cloud.r-project.org")
}

## mlbench ----
if (require("mlbench")) {
  require("mlbench")
} else {
  install.packages("mlbench", dependencies = TRUE,
                   repos = "https://cloud.r-project.org")
}
# Load required libraries
if (!is.element("ggplot2", installed.packages()[, 1])) {
  install.packages("ggplot2", dependencies = TRUE)
}
if (!is.element("GGally", installed.packages()[, 1])) {
  install.packages("GGally", dependencies = TRUE)
}
if (!is.element("dplyr", installed.packages()[, 1])) {
  install.packages("dplyr", dependencies = TRUE)
}
if (!is.element("reshape2", installed.packages()[, 1])) {
  install.packages("reshape2", dependencies = TRUE)
}

## plumber ----
if (require("plumber")) {
  require("plumber")
} else {
  install.packages("plumber", dependencies = TRUE,
                   repos = "https://cloud.r-project.org")
}
if (require("dplyr")) {
  require("dplyr")
} else {
  install.packages("dplyr", dependencies = TRUE,
                   repos = "https://cloud.r-project.org")
}

```
# Milestone 1 ----

### Issue 1: DESCRIPTIVE STATISTICS ----
```{r}
library(readr)
# Load the mice package
library(mice)


# Read the CSV file with read.csv
library(readr)
breast_cancer <-read_csv("data/breast cancer.csv", col_types = cols(...33 = col_skip()))



# Dimensions
dim(breast_cancer)

# Data Types
sapply(breast_cancer, class)

# Measure of Frequency
breast_cancer_frequencies <- table(breast_cancer$symmetry_se)

# Measure of Central Tendency
breast_cancer_freq <- breast_cancer$symmetry_se
cbind(frequency = table(breast_cancer_freq),
      percentage = prop.table(table(breast_cancer_freq)) * 100)

breast_cancer_mode <- names(table(breast_cancer$smoothness_se))[
  which(table(breast_cancer$smoothness_se) ==
          max(table(breast_cancer$smoothness_se)))
]
print(breast_cancer_mode)

# Measure of Distribution
summary(breast_cancer)

# Measure of Standard deviation
sapply(breast_cancer[, c(3:32)], sd)

# Measure the variance of each variable
sapply(breast_cancer[, c(3:32)], var)

# Measure of kurtosis of each variable
if (!is.element("e1071", installed.packages()[, 1])) {
  install.packages("e1071", dependencies = TRUE)
}
require("e1071")

sapply(breast_cancer[, 3:32], kurtosis, type = 2)

# Measure of skewness of each variable
sapply(breast_cancer[, 3:32], skewness, type = 2)



### Issue 2: STATISTICAL TEST (ANOVA)----
breast_cancer_one_way_anova <- aov(compactness_mean ~ diagnosis, data = breast_cancer)
summary(breast_cancer_one_way_anova)

### Issue 3: UNIVARIATE AND MULTIVARIATE PLOTS----
# Load required libraries
library(ggplot2)
library(GGally)
library(dplyr)
library(reshape2)
str(breast_cancer)

# Select only the numeric columns and the target variable
numeric_columns <- breast_cancer[, 3:32]
target_variable <- breast_cancer[, 2]

# Combine the numeric columns and the target variable into one data frame
data <- cbind(numeric_columns, Diagnosis = target_variable)

# Convert the target variable to a factor
data$Diagnosis <- as.factor(data$diagnosis)

# Univariate Plots: Density Plots
# Melt the data for ggplot
melted_data <- melt(data, id.vars = "diagnosis")
# Create density plots for each variable
ggplot(melted_data, aes(x = value, fill = diagnosis)) + 
  geom_density(alpha = 0.5) + 
  facet_wrap(~variable, scales = "free") + 
  theme_minimal() +
  labs(title = "Density Plots of Each Variable", x = "Value", y = "Density")

# Multivariate Plots: Pair Plot with GGally
ggpairs(data, columns = 1:30, aes(color = diagnosis), 
        upper = list(continuous = "points"),
        lower = list(continuous = "smooth"),
        diag = list(continuous = "densityDiag")) + 
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

# Correlation Heatmap with ggplot2
# Compute the correlation matrix
cor_matrix <- cor(numeric_columns)

# Melt the correlation matrix for ggplot2
melted_cor_matrix <- melt(cor_matrix)

# Create a heatmap
ggplot(data = melted_cor_matrix, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 0, limit = c(-1, 1), space = "Lab", 
                       name = "Correlation") +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, 
                                   size = 12, hjust = 1)) +
  coord_fixed() +
  labs(title = "Correlation Heatmap")
```
# Milestone 2 ----
### Issue 5: MISSING VALUES, DATA IMPUTATION, AND DATA TRANSFORMATION ----

```{r}
# Is there missing data?
any(is.na(breast_cancer))

```
# Milestone 3 ----
### Issue 6: TRAINING THE MODEL MILESTONE ----

```{r}
# split the dataset
train_index <- createDataPartition(breast_cancer$diagnosis,
                                   p = 0.75,
                                   list = FALSE)
breast_cancer_train <- breast_cancer[train_index, ]
breast_cancer_test <- breast_cancer[-train_index, ]
# Remove rows with any missing values
breast_cancer_train <- na.omit(breast_cancer_train)
# Check the data type
str(breast_cancer_train$diagnosis)

# Convert to factor if needed
breast_cancer_train$diagnosis <- as.factor(breast_cancer_train$diagnosis)



### Classification: LDA with k-fold Cross Validation ----
train_control <- trainControl(method = "cv", number = 5)

breast_cancer_model_lda <-
  caret::train(`diagnosis` ~ ., data = breast_cancer_train,
               trControl = train_control, na.action = na.omit, method = "lda2",
               metric = "Accuracy")

### Cross-validation ----
# check if the outcome variable is a factor
str(breast_cancer$diagnosis)



# Set the seed for reproducibility
set.seed(123)

# Define the training control
library(caret)
# Define the training control
train_control <- trainControl(method = "cv",
                              number = 10,
                              search = "grid",
                              classProbs = TRUE,
                              summaryFunction = multiClassSummary)

# Model training Random Forest
### Train a Random Forest model----
library(randomForest)
# Convert the class levels to valid variable names
levels(breast_cancer_train$diagnosis) <- make.names(levels(breast_cancer_train$diagnosis))

# Train the Random Forest model
rf_model <- train(diagnosis ~ ., data = breast_cancer_train, method = "rf", trControl = train_control)

rf_model <- train(diagnosis ~ ., data = breast_cancer_train, method = "rf", trControl = train_control)

# Make predictions on the testing set
predictions <- predict(rf_model, newdata = breast_cancer_test)

### Train an SVM model----
library(e1071)
svm_model <- train(diagnosis ~ ., data = breast_cancer_train, method = "svmRadial", trControl = train_control)

# Model sampling and comparison
# Define the control parameters for resampling
train_control <- trainControl(method = "cv", number = 10)  # 10-fold cross-validation

# List of trained models
models_list <- list(
  Random_Forest = rf_model,
  SVM = svm_model
)

# Compare model performance using resampling techniques
results <- resamples(models_list, control = train_control)

# Summarize and compare model performance metrics (e.g., accuracy, sensitivity, specificity)
summary(results)
```
# Milestone 4 ----
### Issue 7: HYPER-PARAMETER TUNING AND ENSEMBLES
```{r}
# Load necessary libraries
library(caret)
library(randomForest)

# Define the grid of hyperparameters for mtry
grid <- expand.grid(mtry = c(2, 4, 6, 8, 10)) # Vary the number of variables randomly sampled as candidates at each split

# Set up the control parameters for grid search
control <- trainControl(method = "cv", number = 5)

# Perform grid search for hyperparameters (only mtry)
model_grid_search <- train(symmetry_se ~ ., data = breast_cancer_train, method = "rf",
                           trControl = control, tuneGrid = grid)
```
# Milestone 5 ----
```{r}
### Issue 8: CONSOLIDATION ---
saveRDS(breast_cancer_model_lda, "./models/saved_breast_cancer_model_lda.rds")
```
